\vspace{-\baselineskip} Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-2\baselineskip}

### 2017-06-12 paper/ms.tex
edits to appendix

To start, we define a rank-$R$ semiseparable matrix as a matrix $K$ where the
elements can be written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.  An example of this type of matrix is given in appendix \ref{operations}.
In this representation, rather than storing the full $N \times N$ matrix $K$,
it is sufficient to only store the $N \times R$ matrices $U$ and $V$, and the
$N$ diagonal entries of $A$, for a total storage volume of $(2\,R+1)\,N$
floating point numbers.
Given a semiseparable matrix of this form, it is known that matrix-vector
products and matrix-matrix products can be evaluated in $\mathcal{O}(N)$
operations and the interested reader is directed to \citet{Vandebril:2007} for
more details about semiseparable matrices in general.

\subsection{Cholesky factorization of positive definite semiseparable matrices}

@@

\section{Semiseparable operations} \label{operations}

To help clarify equations \eq{system}, we write out here an example of a semiseparable matrix with a
single \celeriteterm\ term with $c=d=1$:
\begin{eqnarray}\eqlabel{semi-separable-structure}
K&=&%
\begin{pmatrix}
    {\scriptscriptstyle a+{\sigma_1}^2 }& {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)} & {\scriptscriptstyle e^{-\tau_{31}}\left(a\,\cos{\tau_{31}}+b\,\sin{\tau_{31}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{21}}\left(a\,\cos{\tau_{21}}+b\,\sin{\tau_{21}}\right)}&{\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \ldots\\
    {\scriptscriptstyle e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)}&{\scriptscriptstyle  e^{-\tau_{32}}\left(a\,\cos{\tau_{32}}+b\,\sin{\tau_{32}}\right)} & \scriptscriptstyle{a + {\sigma_3}^2} & \ldots \\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&=&
\begin{pmatrix}
    {\scriptscriptstyle a+{\sigma_1}^2} & {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle a+{\sigma_2}^2 }& {\scriptscriptstyle 0} & \ldots\\
    {\scriptscriptstyle 0} & {\scriptscriptstyle 0} & {\scriptscriptstyle a + {\sigma_3}^2} & \ldots\\
    \ldots & \ldots & \ldots & \ldots
\end{pmatrix}\\
&+&\mathrm{tril}\left[\begin{pmatrix}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix}
\times
\begin{pmatrix}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix}^T\right]\\
&+&
\mathrm{triu}\left[\begin{pmatrix}
 {\scriptscriptstyle e^{t_1} \cos{t_1}} & {\scriptscriptstyle e^{t_1} \sin{t_1}} \\
 {\scriptscriptstyle e^{t_2} \cos{t_2}} & {\scriptscriptstyle e^{t_2} \sin{t_2}}\\
 {\scriptscriptstyle e^{t_3} \cos{t_3}} & {\scriptscriptstyle e^{t_3} \sin{t_3}}\\
 \ldots & \ldots
\end{pmatrix}
\times
\begin{pmatrix}
    {\scriptscriptstyle e^{-t_1}\left(a\,\cos{t_1}+b\,\sin{t_1}\right)}&
    {\scriptscriptstyle e^{-t_1}\left(a\,\sin{t_1}-b\,\cos{t_1}\right)}\\
    {\scriptscriptstyle e^{-t_2}\left(a\,\cos{t_2}+b\,\sin{t_2}\right)}&
    {\scriptscriptstyle e^{-t_2}\left(a\,\sin{t_2}-b\,\cos{t_2}\right)}\\
    {\scriptscriptstyle e^{-t_3}\left(a\,\cos{t_3}+b\,\sin{t_3}\right)}&
    {\scriptscriptstyle e^{-t_3}\left(a\,\sin{t_3}-b\,\cos{t_3}\right)}\\
   \ldots & \ldots
\end{pmatrix}^T\right]\\
&=& A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T).
\end{eqnarray}

In addition to solving $K^{-1} {\bf y}$, several other operations can be
carried out recursively in $\mathcal{O}(N)$, simulation, multiplication,
and prediction, which we describe in the following sections.

@@

Multiplication of the full covariance matrix times a vector can be accomplished in $\mathcal{O}(RN)$
operations with a stable recursive algorithm using the semiseparable form:
\begin{equation}
\bvec{y} =  \left(A + \mathrm{tril}(U V^T) + \mathrm{triu}(V U^T)\right) \bvec{z}.
\end{equation}
The algorithm for multiplication is given by
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j} \left(f^+_{n+1,j} + \tilde U_{n+1,j} z_{n+1}\right)\\
f^-_{n,j} &=& \phi_{n,j} \left(f^-_{n-1,j} + \tilde V_{n-1,j} z_{n-1}\right)\\
y_n &=& A_{n,n} z_n + \sum_{j=1}^R \left[ \tilde V_{n,j} f^+_{n,j} + \tilde U_{n,j} f^-_{n,j}\right],
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$ for all $j=1,...,R$.  The recursion requires
two sweeps, a sweep downward from $n=N-1$ to $n=1$ to compute $f^+_{n,j}$ and a sweep upwards
from $n=2$ to $n=N$ to compute $f^-_{n,j}$.

Multiplication of the full covariance matrix is useful, for example, for isolating
a particular component of a kernel, which amounts to a customized filter of a dataset.
For example, a white-noise filter is given by:
\begin{eqnarray}
\bar f &=& K(\bvec{t},\bvec{t}) \left[K(\bvec{t},\bvec{t})+\sigma^2I\right]^{-1} \bvec{y},\\
&=& K(\bvec{t},\bvec{t}) \bvec{z}
\end{eqnarray}
where $\bvec{y}$ is the vector of measurements and $\bvec{z}$ may be computed from the Cholesky
decomposition as described above.  The second step multiplication of the white-noise-free
kernel times $\bvec{z}$ may be accomplished in $\mathcal{O}(RN)$ steps with this multiplication algorithm.
Note that for this particular example it is more efficient to compute $\bar f = \bvec{y} -\sigma^2 I \bvec{z}$,
but in general the filtering kernel will not differ from the full kernel by a diagonal matrix.

@@

%The variance evaluated at future predicted times $\bvec{t^*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
%\begin{eqnarray}
%\mathrm{var}(f^*_{i+1}) &=& K(t^*_{i+1},t^*_{i+1}) - K(t^*_{i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t^*_{i+1})\\
%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} \\
%&\times& e^{-\beta_p(t^*_{i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}\\
%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} P_{i,p,q}\\
%\end{eqnarray}
%for $1 \le p,q \le J$,
%where
%\begin{equation}
%P_{i,p,q} = e^{-\beta_p(t^*_{i}-\bvec{t}^T)}
%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}.
%\end{equation}
%For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
%with the Cholesky decomposition.
%Subsequent $P_{i+1,p,q}$ may be obtained recursively as
%\begin{equation}
%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t^*_{i+1}- t^*_{i})} P_{i,p,q},
%\end{equation}
%in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
%Note that this variance formula {\it only} works for prediction; we have
%yet to derive a formula for which $\bvec{t^*}$ and $\bvec{t}$ are
%interspersed.

>>>>

To start, we define a rank-$R$ semiseparable matrix as a matrix $K$ where the
elements can be written as
\begin{eqnarray}
K_{n,m} = \left\{\begin{array}{ll}
    \sum_{r=1}^R U_{n,r}\,V_{m,r} & \mathrm{if}\,m<n \\
    A_{n,n} & \mathrm{if}\,m=n \\
    \sum_{r=1}^R U_{m,r}\,V_{n,r} & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
for some $N \times R$-dimensional matrices $U$ and $V$, and an $N \times
N$-dimensional diagonal matrix $A$.
This definition can be equivalently written as
\begin{eqnarray}\eqlabel{semi-sep}
K = A + \mathrm{tril}(U\,V^\T) + \mathrm{triu}(V\,U^\T)
\end{eqnarray}
where the $\mathrm{tril}$ function extracts the strict lower triangular
component of its argument and $\mathrm{triu}$ is the equivalent strict upper
triangular operation.
In this representation, rather than storing the full $N \times N$ matrix $K$,
it is sufficient to only store the $N \times R$ matrices $U$ and $V$, and the
$N$ diagonal entries of $A$, for a total storage volume of $(2\,R+1)\,N$
floating point numbers.
Given a semiseparable matrix of this form, it is known that matrix-vector
products and matrix-matrix products can be evaluated in $\mathcal{O}(N)$
operations and the interested reader is directed to \citet{Vandebril:2007} for
more details about semiseparable matrices in general.

\subsection{Cholesky factorization of positive definite semiseparable matrices}
\sectlabel{cholesky}

@@

\response{
\section{Semiseparable matrix operations} \sectlabel{operations}

In this appendix, we summarize some algorithms for manipulating semiseparable
matrices and discuss how these can be used with GP models and \celerite.

The product of a rank-$R$ semiseparable matrix with general vectors and
matrices can be computed in $\mathcal{O}(N\,R)$ operations
\citep{Vandebril:2007}.
In the context of \celeriteterm\ models, na\"ive application of these methods
will result in numerical overflow and underflow issues much like we found with
the Cholesky factorization.
To avoid these errors, we derive the following numerically stable algorithm
for computing the dot product of a \celeriteterm\ covariance matrix
\begin{eqnarray}
\bvec{y} = K\,\bvec{z}
\end{eqnarray}
using the semiseparable representation for $K$ defined in \eqalt{reparam}:
\begin{eqnarray}
f^+_{n,j} &=& \phi_{n+1,j}\,\left(f^+_{n+1,j} + \tilde{U}_{n,j}\,z_{n+1}\right) \\
f^-_{n,j} &=& \phi_{n,j}\,\left(f^-_{n-1,j} + \tilde{V}_{n-1,j}\,z_{n-1}\right) \\
y_n &=& A_{n,n}\,z_n + \sum_{j=1}^{2\,J} \left[ \tilde{V}_{n,j} f^+_{n,j} +
    \tilde U_{n-1,j} f^-_{n,j}\right] \quad,
\end{eqnarray}
where $f^+_{N,j} = 0$ and $f^-_{1,j} = 0$.
This algorithm requires two sweeps~--~$n=1, \ldots, N$ to compute $f^-$ and
$n=N,\ldots,1$ to compute $f^+$~--~with a scaling of $\mathcal{O}(N\,J)$.

@@

@@

%The variance evaluated at future predicted times $\bvec{t^*} > \bvec{t}$ may be obtained in a similar manner (equation 2.19 from Rasmussen \& Williams):
%\begin{eqnarray}
%\mathrm{var}(f^*_{i+1}) &=& K(t^*_{i+1},t^*_{i+1}) - K(t^*_{i+1},\bvec{t}^T)K(\bvec{t},\bvec{t})^{-1} K(\bvec{t},t^*_{i+1})\\
%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_p\sum_q \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} \\
%&\times& e^{-\beta_p(t^*_{i}-\bvec{t}^T)}K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}\\
%&=& K(t^*_{i+1},t^*_{i+1}) - \sum_{p,q} \alpha_p \alpha_q e^{-(\beta_p+\beta_q)(t^*_{i+1}-t^*_{i})} P_{i,p,q}\\
%\end{eqnarray}
%for $1 \le p,q \le J$,
%where
%\begin{equation}
%P_{i,p,q} = e^{-\beta_p(t^*_{i}-\bvec{t}^T)}
%K(\bvec{t},\bvec{t})^{-1} e^{-\beta_q(t^*_{i}-\bvec{t})}.
%\end{equation}
%For $i=1$, $P_{1,p,q}$ can be obtained in $\mathcal{O}(J^2N)$ operations
%with the Cholesky decomposition.
%Subsequent $P_{i+1,p,q}$ may be obtained recursively as
%\begin{equation}
%P_{i+1,p,q} = e^{-(\beta_p+\beta_q)(t^*_{i+1}- t^*_{i})} P_{i,p,q},
%\end{equation}
%in $\mathcal{O}(J^2M)$ operations, for a total of $\approx \mathcal{O}(J^2(N+M))$ operations.
%Note that this variance formula {\it only} works for prediction; we have
%yet to derive a formula for which $\bvec{t^*}$ and $\bvec{t}$ are
%interspersed.
}
